Design and Implementation of Probabilistic Programming Language Anglican

### 2016-11-27 paper/paper.tex
update

\texttt{anglican.runtime} is the Anglican runtime library. For
convenience, it exposes common mathematical functions
(\texttt{abs}, \texttt{floor}, \texttt{sin}, \texttt{log},
\texttt{exp}, etc.), but most importantly, it provides
definitions of common distributions. Each distribution object
implements the \texttt{anglican.{\linebreak[0]}runtime/{\linebreak[0]}distribution} protocol,
with two methods: \texttt{sample*} and \texttt{observe*}.
The \texttt{sample*} \textit{method} returns a random sample and
roughly corresponds to the \texttt{sample} checkpoint,
the \texttt{observe*} \textit{method} returns the log probability
of the value and roughly corresponds to the \texttt{observe}
checkpoint. The methods can be, and sometimes are called from
handlers of the corresponding checkpoints, but do not have to
be. For example, in LMH either the \texttt{sample*} or
the \texttt{observe*} \textit{method} is called for a
\texttt{sample} checkpoint, depending on whether the value is
drawn or reused.

Macro \texttt{defdist} should be used to define distributions.
\texttt{defdist} takes care of defining a separate type
for every distribution so that multimethods can be dispatched on
distribution types when needed, e.g. for custom proposals. The
Bernoulli distribution could be defined as follows:
\begin{lstlisting}[style=default]
(defdist bernoulli
  "Bernoulli distribution"
  [p]
  (sample* [this] (if (< (rand) p) 1 0))
  (observe* [this value]
    (Math/log (case value
                1 p
                0 (- 1. p)
                0.))))
\end{lstlisting}

Similar to distributions, \textit{random processes}, related to
so called `exchangeable random procedures'\footnote{However, random sequences
generated by Anglican random processes are not required to be
exchangeable.}, are defined using
macro \texttt{defproc} and implement the
\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol. The protocol
has two methods --- \texttt{produce}, which returns a
distribution object, and \texttt{absorb}, which returns a new
random process object obtained by absorbing a value drawn from
the process. Here is a definition of the Poisson process:
\begin{lstlisting}[style=default]
(defproc PP
  "Poisson point process"
  [rate] [dist (exponential rate)]
  (produce [this] dist)
  (absorb [this value] this))
\end{lstlisting}
While distributions are random and non-functional,
random processes are deterministic and functional, hence
\texttt{produce} and \texttt{absorb} are called directly and do
not have corresponding special forms in Anglican:
\begin{itemize}
    \item Values produced by the process are sampled from the distribution
        returned by \texttt{produce}.
    \item When a value is observed from the distribution, a new
        process, with the observed value absorbed, is returned
        by \texttt{absorb}.
\end{itemize}
A common pattern of programming with processes in Anglican is
recursively computing a new process based on the observed value:
\begin{lstlisting}[style=default]
(loop [process (HawkesPointProcess)
       [value & values] values]
  (let [dist (produce process)]
    (observe dist value)
    (recur (absorb process value) values)))
\end{lstlisting}
Unlike conventional implementations of random processes,
Anglican's random processes do not have mutable state.

>>>>

\texttt{anglican.runtime} is the Anglican runtime library. For
convenience, it exposes common mathematical functions
(\texttt{abs}, \texttt{floor}, \texttt{sin}, \texttt{log},
\texttt{exp}, etc.), but most importantly, it provides
definitions of common distributions. Each distribution object
implements a distribution interface with the \texttt{sample*} and \texttt{observe*}
methods; this interface is defined using Clojure's protocol mechanism
(\texttt{anglican.{\linebreak[0]}runtime/{\linebreak[0]}distribution}).
The \texttt{sample*} \textit{method} returns a random sample and
roughly corresponds to the \texttt{sample} checkpoint,
the \texttt{observe*} \textit{method} returns the log probability
of the value and roughly corresponds to the \texttt{observe}
checkpoint. The methods can be, and sometimes are called from
handlers of the corresponding checkpoints, but do not have to
be. For example, in LMH either the \texttt{sample*} or
the \texttt{observe*} \textit{method} is called for a
\texttt{sample} checkpoint, depending on whether the value is
drawn or reused.

Macro \texttt{defdist} should be used to define distributions.
\texttt{defdist} takes care of defining a separate type
for every distribution so that Clojure multimethods (or overloaded
methods) can be dispatched on
distribution types when needed, e.g. for custom proposal distributions 
used in an inference algorithm. The
Bernoulli distribution could be defined as follows:
\begin{lstlisting}[style=default]
(defdist bernoulli
  "Bernoulli distribution"
  [p]
  (sample* [this] (if (< (rand) p) 1 0))
  (observe* [this value]
    (Math/log (case value
                1 p
                0 (- 1. p)
                0.))))
\end{lstlisting}

Similar to distributions, \textit{random processes}, related to
so called `exchangeable random procedures'\footnote{However, random sequences
generated by Anglican random processes are not required to be
exchangeable.}, are defined using
macro \texttt{defproc} and implement the
\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol (that is,
an interface for these random-sequence objects). The protocol
has two methods --- \texttt{produce}, which returns a
distribution object, and \texttt{absorb}, which returns a new
random process object obtained by absorbing a value drawn from
the process. Here is a definition of the Poisson process:
\begin{lstlisting}[style=default]
(defproc PP
  "Poisson point process"
  [rate] [dist (exponential rate)]
  (produce [this] dist)
  (absorb [this value] this))
\end{lstlisting}
While distributions are random and non-functional,
random processes are deterministic and functional, hence
\texttt{produce} and \texttt{absorb} are called directly and do
not have corresponding special forms in Anglican:
\begin{itemize}
    \item Values produced by the process are sampled from the distribution
        returned by \texttt{produce}.
    \item When a value is observed from the distribution, a new
        process, with the observed value absorbed, is returned
        by \texttt{absorb}.
\end{itemize}
A common pattern of programming with processes in Anglican is
recursively computing a new process based on the observed value:
\begin{lstlisting}[style=default]
(loop [process (HawkesPointProcess)
       [value0 & values0] values] 
  (let [dist (produce process)]
    (observe dist value0)
    (recur (absorb process value0) values0)))
\end{lstlisting}
Unlike conventional implementations of random processes,
Anglican's random processes do not have mutable state.
